---
title: "p8105_hw2_hq2229"
author: "Hantang Qin"
output: github_document
date: "2025-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Problem1 

Clean csv
```{r}

# ---- P1: libraries ----
library(tidyverse)
library(janitor)
library(lubridate)

# ---- 1A. pols-month.csv ----
pols <- read_csv("fivethirtyeight_datasets/pols-month.csv", show_col_types = FALSE) %>%
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) %>%
  mutate(
    month    = month.name[month_num],
    president = if_else(prez_gop == 1, "gop", "dem")
  ) %>%
  select(year, month, gov_gop, gov_dem, sen_gop, sen_dem, rep_gop, rep_dem, president)

# ---- 1B. snp.csv ----
snp <- read_csv("fivethirtyeight_datasets/snp.csv", show_col_types = FALSE) %>%
  mutate(date = mdy(date)) %>%                     # parse m/d/Y
  transmute(
    year  = year(date),
    month = as.character(month(date, label = TRUE, abbr = TRUE)),   # "Jan", "Feb", ...
    month = month.name[match(month, month.abb)],                    # -> "January" etc.
    close
  ) %>%
  arrange(year, match(month, month.name))

# ---- 1C. unemployment.csv ----
unemp <- read_csv("fivethirtyeight_datasets/unemployment.csv", show_col_types = FALSE) %>%
  pivot_longer(
    cols = Jan:Dec,
    names_to = "month_abbr",
    values_to = "unemployment_rate"
  ) %>%
  mutate(
    month = month.name[match(month_abbr, month.abb)],
    year  = as.integer(Year)
  ) %>%
  select(year, month, unemployment_rate)

# ---- 1D. Join all three ----
final_data <- pols %>%
  left_join(snp,   by = c("year", "month")) %>%
  left_join(unemp, by = c("year", "month")) %>%
  arrange(year, match(month, month.name))

# ---- Quick descriptors for inline text ----
p1_n_obs   <- nrow(final_data)
p1_n_vars  <- ncol(final_data)
p1_yearmin <- min(final_data$year, na.rm = TRUE)
p1_yearmax <- max(final_data$year, na.rm = TRUE)


```

After cleaning and merging, the final monthly panel has **`r p1_n_obs` observations** and **`r p1_n_vars` variables**, spanning **`r p1_yearmin`–`r p1_yearmax`**.  
Key variables include political control (`president`, plus `gov_*`, `sen_*`, `rep_*`), the **S&P closing value** (`close`), and the **unemployment rate** (`unemployment_rate`).  
Putting everything in one tidy dataset makes it easy to line up political changes with market and labor indicators over time.

Problem 2
```{r}
### Problem 2 — Trash Wheel (my version, explicit ranges)

library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)

# Relative path (per HW)
trash_xlsx <- "202509 Trash Wheel Collection Data.xlsx"

# Mr. Trash Wheel
Mr_trash_wheel =
  read_excel(
    trash_xlsx,
    sheet = "Mr. Trash Wheel",
    range = "A2:N709"
  ) |>
  janitor::clean_names() |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    type = "Mr. Trash Wheel",
    year = as.character(year)
  ) |>
  select(date, type, everything(), -dumpster)

# Professor Trash Wheel
Professor_trash_wheel =
  read_excel(
    trash_xlsx,
    sheet = "Professor Trash Wheel",
    range = "A2:M134"
  ) |>
  janitor::clean_names() |>
  mutate(
    type = "Professor Trash Wheel",
    year = as.character(year)
  ) |>
  select(date, type, everything(), -dumpster)

# Gwynnda / Gwynns Falls Trash Wheel
# (use the sheet name in your file; if it's "Gwynnda Trash Wheel" instead, change here)
Gwynns_trash_wheel =
  read_excel(
    trash_xlsx,
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351"
  ) |>
  janitor::clean_names() |>
  mutate(
    type = "Gwynnda Trash Wheel",
    year = as.character(year)
  ) |>
  select(date, type, everything(), -dumpster)

# Combine all three
trash_wheel =
  bind_rows(Mr_trash_wheel, Professor_trash_wheel, Gwynns_trash_wheel) |>
  arrange(date)

# --- Required summaries (used in inline R) ---
tw_n_obs      <- nrow(trash_wheel)
tw_n_devices  <- dplyr::n_distinct(trash_wheel$type)

tw_prof_total_tons <-
  Professor_trash_wheel |>
  summarise(total = sum(weight_tons, na.rm = TRUE)) |>
  pull(total) |>
  round(2)

# If the month column is month names, this works; if it's numeric, swap "June" for 6
tw_gwyn_jun22_cigs <-
  Gwynns_trash_wheel |>
  filter(year == "2022", month == "June") |>
  summarise(total = sum(cigarette_butts, na.rm = TRUE)) |>
  pull(total)


```

The cleaned dataset includes **`r tw_n_obs` observations** from **`r tw_n_devices` trash wheels**: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda Trash Wheel. Each entry records the device `type`, the `date`, and measures of trash collected, including `weight_tons`, `volume_cubic_yards`, and counts such as `plastic_bottles`, `polystyrene`, and `cigarette_butts`.  In terms of totals, **Professor Trash Wheel** has collected **`r tw_prof_total_tons` tons** of trash overall. Meanwhile, **Gwynnda** picked up **`r format(tw_gwyn_jun22_cigs, big.mark = ",")` cigarette butts** just in June 2022.  






Problem 3
```{r}
### Problem 3 — Zillow Rent Index (NYC)

library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
library(knitr)

# Borough map: county -> borough
borough_map <- c(
  "New York" = "Manhattan",
  "Kings"    = "Brooklyn",
  "Queens"   = "Queens",
  "Bronx"    = "Bronx",
  "Richmond" = "Staten Island"
)

# ---------------------------
# 1) ZIP reference
# ---------------------------
zip_ref <-
  read_csv("zillow_data/Zip Codes.csv", na = c("NA", ".", ""), show_col_types = FALSE) |>
  clean_names() |>
  mutate(
    zip_code = as.character(zip_code),
    county   = as.character(county)
  )

# ---------------------------
# 2) ZORI wide data (2015-01-31 .. 2024-08-31)
# ---------------------------
zori_raw <-
  read_csv("zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv",
           na = c("NA", ".", ""), show_col_types = FALSE) |>
  clean_names() |>
  mutate(
    county_name = str_remove(county_name, " County$")
  ) |>
  # remove the leading "x" on date columns
  rename_with(~ str_remove(.x, "^x"), .cols = starts_with("x")) |>
  rename(zip_code = region_name, county = county_name) |>
  # keep rows that have at least one non-NA price across the date cols
  filter(!if_all(matches("^\\d{4}_\\d{2}_\\d{2}$"), is.na)) |>
  # keep identifiers + all date columns
  select(zip_code, county, any_of(c("region_id","size_rank")), matches("^\\d{4}_\\d{2}_\\d{2}$")) |>
  mutate(
    zip_code = as.character(zip_code),
    county   = as.character(county)
  )

# ---------------------------
# 3) Join + pivot longer  (SAFE SELECT)
# ---------------------------
joint_data <-
  right_join(zip_ref, zori_raw, by = c("zip_code","county")) |>
  pivot_longer(
    cols      = matches("^\\d{4}_\\d{2}_\\d{2}$"),
    names_to  = "date_chr",
    values_to = "price"
  ) |>
  mutate(
    date    = ymd(str_replace_all(date_chr, "_", "-")),
    borough = recode(county, !!!borough_map)
  ) |>
  select(any_of(c(
    "state","city","county","borough","state_fips","county_fips","county_code","metro",
    "neighborhood","zip_code","region_id","size_rank","date","price","file_date"
  ))) |>
  arrange(zip_code, date)

# ---------------------------
# 4) Descriptors (for inline R)
# ---------------------------
p3_n_obs          <- nrow(joint_data)
p3_n_zip          <- n_distinct(joint_data$zip_code)
p3_n_neighborhood <- n_distinct(joint_data$neighborhood)
p3_date_min       <- suppressWarnings(min(joint_data$date, na.rm = TRUE))
p3_date_max       <- suppressWarnings(max(joint_data$date, na.rm = TRUE))

# ---------------------------
# 5) ZIPs present in reference but missing in ZORI
# ---------------------------
zips_missing <- anti_join(
  zip_ref |> select(any_of(c("zip_code","county","neighborhood","city","state"))),
  zori_raw |> select(zip_code, county),
  by = c("zip_code","county")
)
n_missing_zips   <- nrow(zips_missing)
missing_examples <- head(zips_missing$zip_code, 5)

# ---------------------------
# 6) Largest drop: Jan 2020 → Jan 2021 (drop = 2020 − 2021)
# ---------------------------
drops_20_21 <-
  zori_raw |>
  transmute(
    zip_code, county,
    drop_2020_to_2021 = `2020_01_31` - `2021_01_31`
  ) |>
  filter(!is.na(drop_2020_to_2021)) |>
  left_join(zip_ref |> select(zip_code, any_of("neighborhood")), by = "zip_code") |>
  mutate(borough = recode(county, !!!borough_map))

top10_largest_drop <-
  drops_20_21 |>
  arrange(desc(drop_2020_to_2021)) |>
  select(zip_code, borough, neighborhood, county, drop_2020_to_2021) |>
  slice_head(n = 10)

# ---------------------------
# 7) Largest overall range across full period
# ---------------------------
top10_range <-
  joint_data |>
  group_by(zip_code) |>
  filter(any(!is.na(price))) |>
  summarise(
    price_max   = max(price, na.rm = TRUE),
    price_min   = min(price, na.rm = TRUE),
    price_range = price_max - price_min,
    .groups = "drop"
  ) |>
  arrange(desc(price_range)) |>
  slice_head(n = 10) |>
  left_join(zip_ref |> select(zip_code, any_of(c("neighborhood","county"))), by = "zip_code") |>
  mutate(borough = recode(county, !!!borough_map)) |>
  relocate(borough, neighborhood, county, .after = zip_code)

# ---------------------------
# 8) Output tables
# ---------------------------
kable(top10_largest_drop,
     caption = "Top 10 ZIP codes by largest ZORI drop (Jan 2020 → Jan 2021; drop = 2020 − 2021)")

kable(zips_missing |> slice_head(n = 10),
     caption = "Examples: ZIPs present in ZIP file but missing from ZORI dataset")

kable(top10_range,
     caption = "Top 10 ZIP codes by overall ZORI range")

```
The final tidy dataset has **`r p3_n_obs` observations** across **`r p3_n_zip` ZIP codes**
and **`r p3_n_neighborhood` neighborhoods**, spanning **`r p3_date_min`** to **`r p3_date_max`**.
There are **`r n_missing_zips` ZIP codes** in the reference that don’t appear in ZORI
(e.g., `r paste(missing_examples, collapse = ", ")`), likely PO-box or non-residential areas.
The largest pandemic-era drop (Jan 2020 → Jan 2021) and the widest overall range are shown
in the tables above.


For the change between January 2020 and January 2021, the sharpest drop in rental prices was about 912.5966 dollars in Lower Manhattan. Looking over the entire time period, the largest overall decline was close to 3001.010 dollars, also in Lower Manhattan. Both patterns show how dramatically the COVID-19 pandemic disrupted the rental market.

